Automated assessment of residual plots with computer vision models

### 2024-05-13 paper/paper.tex
up to examples part

\subsection{Model Formulas}\label{model-formulas}

There are various configurations of the computer vision model that can
be used to assess residual plots. We discuss these configurations below
focusing on two key components of the model formula: the input and the
output format.

@@

\caption{Diagram of the architecture of the optimized computer vision model.}\label{fig:unnamed-chunk-2}
\end{figure}

The architecture of the computer vision model is adapted from a
well-established architecture known as VGG16, which has demonstrated
high performance in image classification \citep{simonyan2014very}.

@@

Based on the tuning process described in Section
\ref{training-and-hyperparameter-tuning}, the optimized hyperparameter
values are presented in Table \ref{tab:best-hyperparameter}. It was
observable that a minimum of \(32\) base filters was necessary, with the
preferable choice being \(64\) base filters for both the
\(64 \times 64\) and \(128 \times 128\) models, mirroring the original
VGG16 architecture. The optimized dropout rate for convolutional blocks
hovered around \(0.4\), and incorporating batch normalization for
convolutional blocks proved beneficial for performance.

@@

\subsection{Evaluation Methods}\label{evaluation-methods}

This section remains incomplete as we may modify the ``Results'' section
later on.

@@

understand what the convolutional layer is actually doing (read vgg16
doc)

\subsection{Summary of model
performance}\label{summary-of-model-performance}

@@

The figure highlights that most under-predictions occurred when
\(\hat{D} < 3\), while over-predictions occurred predominantly when
\(3 < \hat{D} < 6\). To comprehend the reasons behind these deviations,
we fitted a regression tree using the \texttt{rpart} R package
\citep{Terry2022rpart} on \(D - \hat{D}\) for all factors used in the
data generating process. The results of the regression tree are provided
in the Appendix.

The regression tree revealed that most issues arose from non-linearity
problems and the presence of a second predictor in the regression model.
When the error distribution had a very small variance, all computer
vision models tended to under-predict the distance. Conversely, when the
error distribution had a large variance, all models tended to
over-predict the distance. Furthermore, for input images representing
null plots, it was expected that the models will over-predict the
distance, as explained in Section \ref{lineup-evaluation}.

Since most of the deviation stemmed from the presence of non-linearity
violations, to further investigate this, we split the test set and
re-evaluated the performance, as detailed in Table
\ref{tab:performance-sub}. It was evident that metrics for null plots
were notably worse compared to other categories. Furthermore, residual
plots solely exhibiting non-normality issues were the easiest to
predict, with very low test root mean square error (RMSE) at around
\(0.3\). Residual plots with non-linearity issues were more challenging
to assess than those with heteroskedasticity or non-normality issues.
Assessing residual plots with heteroskedasticity was not as difficult as
assessing those with non-linearity issues. When multiple violations were
introduced to a residual plot, the performance metrics typically lay
between the metrics for each individual violation.

@@

@@

\subsection{Comparison with human visual inference and conventional
tests}\label{comparison-with-human-visual-inference-and-conventional-tests}

\subsubsection{Overview of the human subject
experiment}\label{overview-of-the-human-subject-experiment}

In order to check the validity of the proposed computer vision model,
residual plots presented in the human subject experiment conducted by
\citet{li2023plot} will be assessed. This study has collected \(7974\)
human responses to \(1152\) lineups. Each lineup contains one randomly
placed true residual plot and 19 null plots. Among the \(1152\) lineups,
\(24\) are attention check lineups in which the visual patterns are
designed to be extremely obvious and very different from the
corresponding to null plots, \(36\) are null lineups where all the
lineups consist of only null plots, \(279\) are lineups with uniform
predictor distribution evaluated by \(11\) participants, and the
remaining \(813\) are lineups with discrete, skewed or normal predictor
distribution evaluated by \(5\) participants. Attention check lineups
and null lineups will not be assessed in the following analysis.

@@

For each lineup used in \citet{li2023plot}, there is one true residual
plot and 19 null plots. While the distance \(D\) for the true residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to predict the approximated distance for both the true
residual plots and the null plots. To have a fair comparison, \(H_0\)
will be rejected if the true residual plot has the greatest approximated
distance among all plots in a lineup. Additionally, the appropriate
conventional tests including the Ramsey Regression Equation
Specification Error Test (RESET) \citep{ramsey1969tests} for
non-linearity and the Breusch-Pagan test \citep{breusch1979simple} for
heteroskedasticity were applied on the same data for comparison.

@@

Table \ref{tab:human-conv-table} provides a summary of the agreement
between decisions made by the computer vision model and conventional
tests. The agreement rates between conventional tests and the computer
vision model are 85.95\% and 79.69\% for residual plots containing
heteroskedasticity and non-linearity patterns, respectively. These
figures are higher than those calculated for visual tests conducted by
human, indicating that the computer vision model exhibits behaviour more
akin to the best available conventional tests. However, Figure
\ref{fig:conv-mosaic} shows that conventional tests tend to reject
residual plots almost exclusively when the computer vision model does,
but they also reject plots when the computer vision model does not. This
suggests that conventional tests are more sensitive than the computer
vision model.

@@

\caption{Rejection rate ($p$-value $\leq0.05$) of conventional test conditional on the computer vision model decision on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. The conventional test rejects more frequently than the computer vision model, and (almost) only rejects when the computer vision model does.}\label{fig:conv-mosaic}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/venn-1}

\caption{Venn diagram for non-rejection decisions made by conventional tests, computer vision model, and visual tests conducted by human.}\label{fig:venn}
\end{figure}

@@

\subsection{Data examples}\label{data-examples}

In this section, we will utilize the trained computer vision model on
three example datasets. These include the dataset associated with the
residual plot displaying a ``left-triangle'' shape, as displayed in
Figure \ref{fig:false-finding}, along with the Boston housing dataset,
and the ``dino'' datasets from the \texttt{datasaurus} R package.
Additionally, we will outline the workflow for employing this model with
the \texttt{autovi} R package.

\subsubsection{Left-triangle}\label{left-triangle}

@@

Figure \ref{fig:false-check} presents the results of the assessment.
Notably, the observed visual signal strength is considerably lower than
the 95\% sample quantile of the null distribution. Moreover, the
bootstrapped distribution suggests that it is highly improbable for the
fitted model to be misspecified as the majority of bootstrapped fitted
models will not be rejected.

Thus, for this particular fitted model, both the visual test and the
computer vision model will not reject \(H_0\). However, the
Breusch-Pagan test will reject \(H_0\) because it can not effectively
utilize information from null plots.

@@

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. On the left, the residual plot exhibiting "left-triangle" shape is displayed. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$.}\label{fig:false-check}
\end{figure}

@@

\subsubsection{Boston housing}\label{boston-housing}

@@

\subsubsection{Datasaurus}\label{datasaurus}

>>>>

\subsection{Model Specifications}\label{model-specifications}

There are various specifications of the computer vision model that can
be used to assess residual plots. We discuss these specifications below
focusing on two key components of the model formula: the input and the
output format.

@@

\caption{Diagram of the architecture of the optimized computer vision model.}\label{fig:cnn-diag}
\end{figure}

The architecture of the computer vision model is adapted from a
well-established architecture known as VGG16, which has demonstrated
high performance in image classification \citep{simonyan2014very}.
Figure \ref{fig:cnn-diag} provides a diagram of the architecture. More
details about the neural network layers used in this study are provided
in the Appendix.

@@

@@

Based on the tuning process described above, the optimized
hyperparameter values are presented in Table
\ref{tab:best-hyperparameter}. It was observable that a minimum of
\(32\) base filters was necessary, with the preferable choice being
\(64\) base filters for both the \(64 \times 64\) and \(128 \times 128\)
models, mirroring the original VGG16 architecture. The optimized dropout
rate for convolutional blocks hovered around \(0.4\), and incorporating
batch normalization for convolutional blocks proved beneficial for
performance.

@@

\subsection{Model Performance}\label{model-performance}

@@

The figure highlights that most under-predictions occurred when
\(\hat{D} < 3\), while over-predictions occurred predominantly when
\(3 < \hat{D} < 6\). For input images representing null plots, it was
expected that the models will over-predict the distance, as explained in
Section \ref{lineup-evaluation}. However, it can not explain the
under-prediction issue. Therefore, we analysed the relationship between
residuals and all the factors involved in the data generating process.
We found that most issues actually arose from non-linearity problems and
the presence of a second predictor in the regression model as
illustrated in Figure \ref{fig:over-under}. When the variance for the
error distribution was small, the optimized model tended to
under-predict the distance. Conversely, when the error distribution had
a large variance, the model tended to over-predict the distance.

Since most of the deviation stemmed from the presence of non-linearity
violations, to further investigate this, we split the test set based on
violation types and re-evaluated the performance, as detailed in Table
\ref{tab:performance-sub}. It was evident that metrics for null plots
were notably worse compared to other categories. Furthermore, residual
plots solely exhibiting non-normality issues were the easiest to
predict, with very low test root mean square error (RMSE) at around
\(0.3\). Residual plots with non-linearity issues were more challenging
to assess than those with heteroskedasticity or non-normality issues.
Assessing residual plots with heteroskedasticity was not as difficult as
assessing those with non-linearity issues. When multiple violations were
introduced to a residual plot, the performance metrics typically lay
between the metrics for each individual violation.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/over-under-1}

\caption{Scatter plots for residuals vs $\sigma$ on test data for the $32 \times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The red lines are smoothing curves produced by fitting gnealized additive models. The area over the zero line in light pink indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
\end{figure}

@@

\subsection{Comparison with Human Visual Inference and Conventional
Tests}\label{comparison-with-human-visual-inference-and-conventional-tests}

\subsubsection{Overview of the Human Subject
experiment}\label{overview-of-the-human-subject-experiment}

In order to check the validity of the proposed computer vision model,
residual plots presented in the human subject experiment conducted by
\citet{li2023plot} will be assessed.

This study has collected \(7974\) human responses to \(1152\) lineups.
Each lineup contains one randomly placed true residual plot and 19 null
plots. Among the \(1152\) lineups, \(24\) are attention check lineups in
which the visual patterns are designed to be extremely obvious and very
different from the corresponding to null plots, \(36\) are null lineups
where all the lineups consist of only null plots, \(279\) are lineups
with uniform predictor distribution evaluated by \(11\) participants,
and the remaining \(813\) are lineups with discrete, skewed or normal
predictor distribution evaluated by \(5\) participants. Attention check
lineups and null lineups will not be assessed in the following analysis.

@@

For each lineup used in \citet{li2023plot}, there is one true residual
plot and 19 null plots. While the distance \(D\) for the true residual
plot depends on the underlying data generating process, the distance
\(D\) for the null plots is zero. We have used our optimized computer
vision model to estimate distance for both the true residual plots and
the null plots. To have a fair comparison, \(H_0\) will be rejected if
the true residual plot has the greatest approximated distance among all
plots in a lineup. Additionally, the appropriate conventional tests
including the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey1969tests} for non-linearity and the Breusch-Pagan
test \citep{breusch1979simple} for heteroskedasticity were applied on
the same data for comparison.

@@

Table \ref{tab:human-conv-table} provides a summary of the agreement
between decisions made by the computer vision model and conventional
tests. The agreement rates between conventional tests and the computer
vision model are 85.95\% and 79.69\% for residual plots containing
heteroskedasticity and non-linearity patterns, respectively. These
figures are higher than those calculated for visual tests conducted by
human, indicating that the computer vision model exhibits behaviour more
akin to the best available conventional tests. However, Figure
\ref{fig:conv-mosaic} shows that the computer vision model does not
always reject when the conventional tests reject. And a small number of
plots will be rejected by computer vision model but not by conventional
tests. This suggests that conventional tests are more sensitive than the
computer vision model.

Figure \ref{fig:pcp} further illustrates the decisions made by visual
tests conducted by human, computer vision models, and conventional
tests, using a parallel coordinate plots. It can be observed that all
three tests will agree with each other for around 50\% of the cases.
When visual tests conducted by human do not reject, there are
substantial amount of cases where computer vision model also not reject
but conventional tests reject. There are much fewer cases that do not
reject by visual tests and conventional tests, but reject by computer
vision models. This indicates computer vision model can behave like
visual tests conducted by human better than conventional tests.
Moreover, there are great proportion of cases where visual tests
conducted by human is the only test who does not reject, .

@@

\caption{Rejection rate ($p$-value $\leq0.05$) of computer vision models conditional on conventional tests on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot.}\label{fig:conv-mosaic}
\end{figure}

@@

@@

\section{Examples}\label{examples}

In this section, we will utilize the trained computer vision model on
three example datasets. These include the dataset associated with the
residual plot displaying a ``left-triangle'' shape, as displayed in
Figure \ref{fig:false-finding}, along with the Boston housing dataset,
and the ``dino'' datasets from the \texttt{datasaurus} R package.

The first example illustrates a scenario where both the computer vision
model and human visual inspection successfully avoid rejecting \(H_0\)
when \(H_0\) is true, contrary to conventional tests. This underscores
the necessity of visually examining the residual plot.

In the second example, we encounter a more pronounced violation of the
model, resulting in rejection of \(H_0\) by all three tests. This
highlights the practicality of the computer vision model, particularly
for less intricate tasks.

The third example presents a situation where the model deviation is
non-typical. Here, the computer vision model and human visual inspection
reject \(H_0\), whereas conventional tests do not. This emphasizes the
benefits of visual inspection and the unique advantage of the computer
vision model, which, like humans, makes decisions based on visual
discoveries.

Additionally, we will outline the workflow for employing this model with
the \texttt{autovi} R package.

\subsection{Left-triangle}\label{left-triangle}

@@

Figure \ref{fig:false-check} presents the results of the assessment.
Notably, the observed visual signal strength is considerably lower than
the 95\% sample quantile of the null distribution. Moreover, the
bootstrapped distribution suggests that it is highly improbable for the
fitted model to be misspecified as the majority of bootstrapped fitted
models will not be rejected. Thus, for this particular fitted model,
both the visual test and the computer vision model will not reject
\(H_0\). However, the Breusch-Pagan test will reject \(H_0\) because it
can not effectively utilize information from null plots.

The attention map at plot B of Figure \ref{fig:false-check} suggests
that the estimation is highly influenced by the top-right and
bottom-right part of the residual plot, as it forms two vertices of the
triangular shape. A principle component analysis is also performed on
the output of the global pooling layer of the computer vision model. As
mentioned in \citet{simonyan2014very}, a computer vision model built
upon the convolutional blocks can be viewed as a feature extractor. For
the \(32 \times 32\) model, there are 256 features outputted from the
global pooling layer, which can be further used for different visual
tasks not limited to distance prediction. To see if these features can
be effectively used for distinguishing null plots and true residual
plot, we linearly project them into the PC1 vs PC2 space as shown in
Figure \ref{fig:false-check}. It can be observed that because the
bootstrapped plots are mostly similar to the null plots, the points
drawn in different colours are mixed together. The true residual plot is
also covered by both the cluster of null plots and cluster of
bootstrapped plots. This accurately reflects our understanding of Figure
\ref{fig:false-lineup}.

@@

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting "left-triangle" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distnace for null plots and bootstrapped plots. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$. (D) PC2 vs PC1 plot for features extracted from the global pooling layer of the computer vision model.  }\label{fig:false-check}
\end{figure}

@@

\subsection{Boston housing}\label{boston-housing}

@@

\subsection{Datasaurus}\label{datasaurus}
